Trade Dependence and Militarized Interstate Disputes

### 2022-12-13 article/article.qmd
Update artilce

```{r libraries}
library(tidyverse)
library(scales)
library(gtsummary)
library(modelsummary)
```

@@

Does trade dependence reduce conflict between states? Debate remains over the role and significance of economic interdependence in promoting peace: different scholars have found that dependence increases, decreases, and has no significant effect on the likelihood of conflict between states. However, this research has generally relied on a broad measure of bilateral trade dependence. This paper will delve deeper into the effect of this economic relationship by exploring product-level dependence. If trade dependence deters conflict between states by increasing the expected cost of conflict through trade disruptions, we should measure the true cost of this disruption. Rarely, if ever, do we see states threaten to or impose total trade embargoes. Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy. Using a more disaggregated measure of trade dependence, I find that this dependence does not have a significant effect on the likelihood of militarized disputes between states.

@@

Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation. Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error. Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP). However, a more disaggregated approach to trade dependence is worth exploring. We are interested in understanding how trade dependence can deter conflict. Rarely (if ever) do states threaten to sever their entire trade relationship with another state. Rather, we regularly see states threaten or enact trade bans on particular products. In fact, states often threaten to inflict the greatest harm on their rival whilst minimizing the costs they will face. A broad understanding of trade dependence does not capture the real threat a potential aggressor faces. A more accurate measure of this factor is product-level trade dependence. Asymmetries at this level have been used as a foreign policy mechanism to deter or punish rival states.

@@

This analysis focuses on all directed state dyads in 2014. This includes `r nrow(full_df) |> comma()` observations. We are interested in whether or not a trade dependent state initiates conflict against another state. This is a directed question. We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on the other.

The narrow timeframe included in this analysis is driven by data constraints. I only have access to complete trade and militarized interstate dispute data for 2014. A natural first step for extending this analysis is including a broader timeframe in the analysis.

@@

The data come from version five of Palmer et al's [-@palmer2022] Militarized Interstate Dispute data collection. This dataset includes all known MIDs from 1816-2014, recorded at the participant level. In 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.

@@

In 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`. The most dependent pairing is between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`: `r pull(trade_dep_max, country1)`'s trade in `r pull(trade_dep_max, cmd_desc_e)` with `r pull(trade_dep_max, country2)` amounts to `r trade_dep_max |> pull(trade_dep_total) |> percent()` of `r pull(trade_dep_max, country1)`'s GDP in 2014.

@@

@@

@@

@@

>>>>

```{r libraries}
library(tidyverse)
library(scales)
library(gtsummary)
```

@@

Does trade dependence reduce conflict between states? Debate remains over the role and significance of economic interdependence in promoting peace: different scholars have found that dependence increases, decreases, and has no significant effect on the likelihood of conflict between states. However, this research has generally relied on a broad measure of bilateral trade dependence. This paper will delve deeper into the effect of this economic relationship by exploring product-level dependence. If trade dependence deters conflict between states by increasing the expected cost of conflict through trade disruptions, we should measure the true cost of this disruption. Rarely, if ever, do we see states threaten to or impose total trade embargoes. Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy. Using this more disaggregated measure of trade dependence and a binary logistic regression model, I find that this dependence does not have a significant effect on the likelihood of militarized disputes between states.

@@

Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation. Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error [@chen2021; @chatagnier2017]. Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP) [@chen2021; @flores-macias2013b; @gartzke2016; @kinne2012a; @lupu2013; @pevehouse2004]. However, a more disaggregated approach to trade dependence is worth exploring. We are interested in understanding how trade dependence can deter conflict. Rarely (if ever) do states threaten to sever their entire trade relationship with another state. Rather, we regularly see states threaten or enact trade bans on particular products. In fact, states often threaten to inflict the greatest harm on their rival whilst minimizing the costs they will face. A broad understanding of trade dependence does not capture the real threat a potential aggressor faces. A more accurate measure of this factor is product-level trade dependence. Asymmetries at this level have been used as a foreign policy mechanism to deter or punish rival states.

@@

This analysis focuses on all directed state dyads in 2013 and 2014. This includes `r nrow(full_df) |> comma()` observations. We are interested in whether or not a trade dependent state initiates conflict against another state. This is a directed question. We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on the other.

The narrow timeframe included in this analysis is driven by data constraints. I only have access to complete trade and militarized interstate dispute data for 2013 and 2014. A natural first step for extending this analysis is including a broader timeframe in the analysis.

@@

The data come from version five of Palmer et al's [-@palmer2022] Militarized Interstate Dispute data collection. This dataset includes all known MIDs from 1816-2014, recorded at the participant level. Between 2013 and 2014, `r full_df |> filter(mid_onset_all == 1) |> nrow() |> comma()` MIDs were initiated.

@@

In 2013 and 2014, product-level trade dependence ranged from `r full_df |> slice_min(trade_dep_total, with_ties = F) |> pull(trade_dep_total) |> percent()` to `r trade_dep_max |> pull(trade_dep_total) |> percent()`, with an average level of trade dependence of `r full_df |> summarise(mean = mean(trade_dep_total)) |> pull(mean) |> percent(accuracy = 0.01)`. The most dependent pairing is between `r pull(trade_dep_max, country1)` and `r pull(trade_dep_max, country2)`: `r pull(trade_dep_max, country1)`'s trade in `r pull(trade_dep_max, cmd_desc_e)` with `r pull(trade_dep_max, country2)` amounts to `r trade_dep_max |> pull(trade_dep_total) |> percent()` of `r pull(trade_dep_max, country1)`'s GDP in 2014.

@@

**See Appendix for Table 1**

@@

**See Appendix for Figure 1**

@@

**See Appendix for Table 2**

## Appendix

### Regression results

@@

### Comparison of product-level and total trade dependence results
